Tree size, microhabitat, and hydraulic traits (interactively) shape drought response in a temperate broadleaf forest

### 2019-08-06 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

### Abstract
Predicting forest responses to drought is an increasingly critical task under climate change effects.
Part of the problem is due the lack of a set definition of drought in recent studies, while often the effects of biophysical and hydraulic traits are interpreted separately.
In this study, we analyze the interaction between these two trait groups using forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
Drought periods were defined by both Palmer Drought Severity Indices (PDSI) and their identification from tree-ring records for 12 species representing 97% of woody productivity.
Each drought scenario (1964-66, 1977, 1999), along with the overall trend, was then tested again leaf hyraulic trait measurements and microhabitat biophysical data.
Individual-level growth responses to three droughts  were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts. 
We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected.

@@

[Understanding forest responses to drought requires increased functional understanding of the factors that confir individual-level vulnerability or resistance.] Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS). Therefore, in order to understand whole-forest response to drought, we need a functional understanding of how responses vary by tree size, microhabitat, and species. There are 3 fundamental questions that must be addressed:

*First, what drives the observed tendency for large trees to suffer more during drought?*  
[@bennett_larger_2015] showed that in forests globally, large trees suffer greater growth reductions during drought. However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms (besides insects) for the observed greater drougth growth reductions of large trees: (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees. Counteracting these effects, (3) the larger root systems of larger trees may confir an advantage in terms of allowing greater access to water, but it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.

@@

*Second, how do species' traits--alone and in interaction with tree size--influence drought response?* 
Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests. The solution is a trait-based approach. Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and SLA ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).

Commonly measured traits including wood density (REFS), leaf mass per area [@abrams_adaptations_1990] [@guerfel_impacts_2009], and ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses and likely correlated with drought resistance in this forest. However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point, which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.

Moreover, there may be an interaction between traits and tree size. It is possible that the pattern observed by [@bennett_larger_2015] could be caused by smaller trees being more drought resistant. Alternatively, larger trees may have more drought-resistant traits as adpatations to greater biophysical challenges.

@@

We need to understand the the factors confirming drought vulnerability or resistance not only for extreme droughts with dramatic impacts on tree growth and mortality, which tend to dominate the literature ([@bennett_larger_2015], REFS), but also for more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade.

@@

**Table 1. Summary of hypotheses, predictions and results** We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction and model dAIC≥2 relative to the approriate null model or to any alternative multivariate model with within 2 dAIC. Parentheses indicate that predictions were partially supported (or rejected)-- i.e., with direction of response matching (or contradicting) the prediction but model dAIC<2 relative to the approriate null model, when response partially matched predicted order for categorical variable (canopy position), or when an alternative multivariate model within 2 dAIC had a contradictory response. "(yes/no) or ("no/yes")" indicates tendancies in opposite directions in univariate tests versus the best full models, respectively. (Table 1.xlsx)

>>>>

### Abstract
Predicting forest responses to drought is an increasingly critical task under climate change effects.
Part of the problem is due to the lack of studies analyzing the confluence of leaf hydraulic traits with biophysical parameters.
In this study, we analyze the interaction between these two trait groups using forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
Drought periods were defined by both Palmer Drought Severity Indices (PDSI) and their identification from tree-ring records for 12 species representing 97% of woody productivity.
Each drought scenario (1964-66, 1977, 1999), along with the overall trend, was then tested against leaf hyraulic trait measurements and microhabitat biophysical data.
Individual-level growth responses to the three individual droughts  were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts. 
We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected.

@@

[Understanding forest responses to drought requires increased functional understanding of the factors that confer individual-level vulnerability or resistance.] Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS). Therefore, in order to understand whole-forest response to drought, we need a functional understanding of how responses vary by tree size, microhabitat, and species. There are 3 fundamental questions that must be addressed:

*First, what drives the observed tendency for large trees to suffer more during drought?*  
@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought. However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms (besides insects) for the observed greater drougth growth reductions of large trees: (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees. Counteracting these effects, (3) the larger root systems of larger trees may confer an advantage in terms of allowing greater access to water, but it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.

@@

*Second, how do species' traits - alone and in interaction with tree size - influence drought response?* 
Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests. The solution is a trait-based approach. Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and specific leaf area (SLA) ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).

Commonly measured traits including wood density (REFS), leaf mass per area [@abrams_adaptations_1990] [@guerfel_impacts_2009], and ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses and likely correlated with drought resistance in this forest. However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication (PLA) and turgor loss point (TLP), which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.

Moreover, there may be an interaction between traits and tree size. It is possible that the pattern observed by @bennett_larger_2015 could be caused by smaller trees being more drought resistant. Alternatively, larger trees may have more drought-resistant traits as adpatations to greater biophysical challenges.

@@

We need to understand the the factors confirming drought vulnerability or resistance not only for extreme droughts with dramatic impacts on tree growth and mortality, which tend to dominate the literature [@bennett_larger_2015] (REFS), but also for more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade.

@@

**Table 1. Summary of hypotheses, predictions and results** We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction and model dAIC>=2 relative to the approriate null model or to any alternative multivariate model with within 2 dAIC. Parentheses indicate that predictions were partially supported (or rejected)-- i.e., with direction of response matching (or contradicting) the prediction but model dAIC<2 relative to the approriate null model, when response partially matched predicted order for categorical variable (canopy position), or when an alternative multivariate model within 2 dAIC had a contradictory response. "(yes/no) or ("no/yes")" indicates tendancies in opposite directions in univariate tests versus the best full models, respectively. (Table 1.xlsx)

```{r, Table 1, eval=TRUE, echo=FALSE, message=FALSE}
library(readxl)
library(kableExtra)
Table1 <- read_xlsx("tables_figures/Table1.xlsx")
Table1 <- sapply(Table1, as.character)
Table1[is.na(Table1)] <- " "

kable(Table1) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```
